Analyzing how sociodemographic characteristics influence voting for the Liberal Party of Canada

### 2024-04-19 paper/paper.qmd
Rename paper (2).qmd to paper.qmd

```{r setup}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(tidyr)
library(dplyr)
library(arrow)
library(ggplot2)
library(kableExtra)
library(bayesplot)
library(rstanarm)
library(modelsummary)
library(lintr)

library(lme4)
```

@@

The analysis reveals age, wealth, and education level have a positive relationship with voting liberal, while children number, gender, and province do not.

@@

>**children_number**: the number of children within the family, which includes only six levels (0,1,2,3,4 and 4+ children)

@@

```
@fig-percent shows the proportion of the population within each subcategory to the population of the entire category after categorizing all voters according to six different demographic categories.

The demographic data indicates that the majority of the population falls within the age range of 60 to 74 years, accounting for approximately 30% of the sample.

Regarding family structure, 40% of individuals do not have children, which includes those who are unmarried, while a two-child family configuration is the next most prevalent, representing 30% of the sample.

In terms of educational attainment, the predominant qualification is a bachelor's degree. High school graduates and individuals without a university degree each make up about 20% of the population.

As for household income, the distribution appears relatively uniform across different brackets, with a notable concentration of individuals earning between 50k to 75k and those earning more than 125k, both groups constituting about 20%.

The gender composition of the population is slightly skewed towards women, who make up just over 50%, while men represent a marginally lower proportion, just above 40%.

Geographically, a significant portion of the electorate, around 50%, resides in Ontario, reflecting a centralization of voters in this province.

@@

```
@fig-vote-percent displays the probability of voters voting for three main parties (the Liberal, the Conservative and the NDP) within each subcategory (e.g. '18-29') under six different demographic categories (e.g. 'age'), after classifying all voters according to these demographic categories.

Age: There appears to be a positive correlation between age and the percentage of votes for the Liberal party. As voters age, from the young demographic of 18-29 to those over 75, support for the Liberals seems to steadily rise from about 50% to just over 50%.

Education Level: From the lowest to the highest education levels, support for the Liberal party shows a relatively stable trend with a slight increase. Those with a bachelor's degree appear to be the most steadfast in their support for the Liberals, with a support rate above 40%.

Gender: Female voters show higher support for the Liberal party than males, indicating that gender may be a significant factor in support for the Liberals.

Number of Children: Families with more children seem to be more inclined to support the Liberal party, especially when there are four or more children, where the support rate peaks at about 45%.

Family Income: The Liberal party enjoys higher support in the middle-income brackets, while there is a decline in support among the higher income bracket (over 125k).

Province: Support for the Liberals is significantly higher in Ontario (ON) than in other provinces, possibly indicating the impact of regional factors on support for the Liberal party.

@@

\begin{align}
y_i | \pi_i &\sim \text{Bern}(\pi_i) \\
\text{logit}(\pi_i) &= \beta_0 + \alpha_{a[i]}^{\text{age}}+ \alpha_{g[i]}^{\text{gender}} + \alpha_{e[i]}^{\text{education level}} 
+\alpha_{f[i]}^{\text{family income}} + \alpha_{c[i]}^{\text{children number}} + \alpha_{s[i]}^{\text{state}} \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\alpha_{a}^{\text{age}} &\sim \text{Normal}(0, 2.5) \text{ for } g = 1,2,3,4,5 \\
\alpha_{g}^{\text{gender}} &\sim \text{Normal}(0, 2.5) \text{ for } g = 1,2 \\
\alpha_{e}^{\text{education level}} &\sim \text{Normal}(0, 2.5) \text{ for } g = 1,2,3,4,5,6 \\
\alpha_{f}^{\text{family income}} &\sim \text{Normal}(0, 2.5) \text{ for } g = 1,2,3,4,5,6 \\
\alpha_{c}^{\text{children number}} &\sim \text{Normal}(0, 2.5) \text{ for } g = 1,2,...,6 \\
\alpha_{s}^{\text{state}} &\sim \text{Normal}(0, \sigma_{\text{state}}^2) \text{ for } s = 1,2,...,9 \\
\sigma_{\text{state}} &\sim \text{Exponential}(1)
\end{align}

>>>>

```{r setup}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(tidyr)
library(dplyr)
library(arrow)
library(ggplot2)
library(kableExtra)
library(bayesplot)
library(rstanarm)
library(modelsummary)
library(lintr)
library(lme4)
```

@@

The statistical analysis indicates that demographic factors such as age, gender, education, family size, and regional differences significantly influence voter preferences towards the Liberal Party. Older age groups and individuals with higher educational qualifications are more likely to support the Liberals, while males and larger families tend to show lower support. Regional variability also plays a crucial role, highlighting the diverse political landscapes across provinces.
Understanding these factors is essential for several reasons. It allows political strategists to craft targeted campaign messages and policies that resonate with specific demographic groups, potentially swaying voter decisions. Additionally, this analysis aids in predicting election outcomes and can serve as a valuable tool for academic and governmental studies aiming to understand and address the diverse needs and concerns of the population. This insight is critical for fostering a more engaged and informed electorate, leading to more representative governance.

@@

- **children_number**: the number of children within the family, which includes only six levels (0,1,2,3,4 and 4+ children)

@@

```
@fig-percent shows the proportion of the population within each subcategory to the population
of the entire category after categorizing all voters according to six different demographic categories.
It reveals that the most populous age segment is 60 to 74 years, comprising roughly
30% of the sample. The 30-44 and 45-59 age groups each make up about 22% of the electorate,
forming the backbone of the voters. The number of elderly voters aged 75 and above is the
smallest. Family structures show that 40% of the cohort is childless, which includes unmarried
individuals, with two-child families being the second most common at 30%.Voters with four
or more children represent the lowest percentage of the electorate, with their combined share
possibly only around 5%. Educationally, bachelor’s degrees prevail as the highest attainment
for approximately 20% of people, closely followed by high school graduates and non-university
educated individuals. Income distribution is fairly even across the spectrum, yet there’s a
discernible cluster of earners between $50k to $75k and above $125k, each group accounting
for 20% of the population. The gender mix leans slightly towards women, who exceed 50%,
with men forming just over 40%. Geographically, half of the voters are centralized in Ontario,
signifying a voter concentration in that province.

@@

```
@fig-vote-percent displays the probability of voters voting for three main parties (the Liberal, the Conservative
and the NDP) within each subcategory (e.g. ‘18-29’) under six different demographic
categories (e.g. ‘age’), after classifying all voters according to these demographic categories.
There appears to be a positive correlation between age and the percentage of votes for the Liberal party. As voters age, support for the Liberal party appears to steadily rise from about
50% among young people aged 18-29 to just over 50% among those over 75. However, this is
slightly less than the support rate for the Conservative party among people aged 45 and older.
Conversely, the NDP sees the opposite trend, with younger voters showing higher support
rates.
From the lowest to the highest education levels, support for the Liberal party shows a relatively
stable trend with a slight increase. Those with a bachelor’s degree appear to be the most
steadfast in their support for the Liberals, with a support rate above 40%.Among voters with
education levels lower than a university certificate, the Conservative party is more popular
than the Liberal party.

Male voters show higher support for the Liberal party than females, but they also tend to favor
the Conservative party more. Female support for the three parties is almost identical. Voters
with one and four children show the highest support for the Liberal party, with 40% of them
voting Liberal. Support for the Liberal party among voters with other numbers of children is
fairly consistent, remaining around 35%. In families with multiple children, support for the
Conservative party significantly increases, surpassing that for the Liberal party. In terms of
income, the Liberal party has higher support among the middle-income class, with a general
trend consistent with the Conservative party. However, support decreases in the higher income
bracket ($100,000 to $125,000), slightly below the Conservative party.
Ontario, the most populous province, shows a slightly higher support rate for the Liberal party
compared to the average support in other provinces, just over 40%, and is slightly ahead of the
other two parties. New Brunswick has the highest support rate for the Liberals, with over half
of the population voting for the Liberal party. Saskatchewan shows the least support for the
Liberals, with only about 20% of the vote. In contrast, the vote shares for the Conservative and
Liberal parties in these two provinces are clearly polarized and significantly different. Overall,
there is a clear preference for the three parties across different provinces, with swing regions
being few.

@@
